Converting Population-Based Simulators to Probabilistic Programs

### 2019-05-16 paper/neurlps2019/main.tex
updated

\section{Introduction}

@@

\item Outline the objectives
\item State the importance of interpretability in the inference of the simulator
\item Our system enables one to understand and interpret the most common paths in a program (simulation).
\item In highlighting this a practitioner can then go back to the field and explore that parameter/s in more depth.
\item This facilitates understanding and the construction of more detailed models and data collection strategies

@@

\bg{The audience is a probabilistic programming audience - so that is what we will target}

@@

This paper first gives an overview of existing Malaria simulators (Section \ref{sec:background}), and proceed by introducing the necessary background on the \textit{pyprob} framework and the concept of universal probabilistic programming (Section \ref{sec:hijackingsimulators}). Our approach is then demonstrated and analysed in the context of a Malaria case study (Section \ref{sec:casestudy}). 
\label{sec:related}

\section{Epidemiology Simulators and Probabilistic Programming}
\label{sec:background}
This will kind of be like a background section.

  \begin{algorithm}
   \caption{How to perform posterior inference over a stochastic simulator with 
   }
    \begin{algorithmic}[1]
      \Function{Merge}{$A,p,q,r$}\Comment{Where A - array, p - left, q - middle, r - right}

        \State replace raw samples statements 
        \State ${n_2} = r - q$
        \State Create Containerized Environment
        \For{$i = 1$ to ${n_1}$}
            \State $L[i] = A[p + i - 1]$
        \EndFor
        \For{$j = 1$ to ${n_2}$}
            \State $R[i] = A[q + j]$
        \EndFor
        \State $L[{n_1} + 1] =  \infty $
        \State $R[{n_2} + 1] =  \infty $
        \State $i = 1$
       \EndFunction

@@

\section{Types of Models and Probabilistic Programming Languages}

To bridge the gap between the statistics and machine learning community to the probabilistic programming community we provide a useful overview to expose our ideas more concretely.

Probabilistic programming languages, broadly speaking\footnote{For a more comprehensive overview see \cite{rainforth2017automating}}, can be classfied into three categories, each type of langauge extends a particular model space, enabling an increasingly complex set models to be represented as programs, but in doing so the notion of posterior and what it means to do inference over the model, at least in a stastically traditional sense, can change.
% \begin{figure}
% \centering
% \includegraphics[]{../models/types_of_pp.png}
% \end{figure}
The first type of langauges are called first-order languages where the priors are well defined and the \textit{joint} program density constructed via the program directly represent the unormalised stastical joint density $p(x,y)$
\section{Methodology}
 TODO

Simulators encode several years, if not decades, of research and development and so by their very nature are structurally complicated. Since the development happens over several decades, usually without documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development. Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular epidemiology simulators. Our framework provides a simple solution and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in\footnote{The framework currently supports 9 popular languages, but there is nothing to stop this being extended to any language of interest.}. In addition to this, we can understand the structure of programs in simulators for both decoding \emph{black-box structures} and for posterior inference. This enables two things. 1) It enables software developers and researchers to understand complicated code bases. 2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to inference outcome. This is critical in several domains, and is particularly important in the medical domain. We provide examples of both in Section~\textbf{TO ADD REF}.

\section{Simulating Diseases}
 \label{sec:background}

@@

In-silico simulators have become a crucial tool in evidence-based decision-making within a large number of disciplines, including statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others.
In many cases, simulation output can augment or even replace real data that may otherwise be costly or even impossible to generate.
Recent advances in hardware have enabled simulations to model increasingly complex systems.
Epidemiology studies the prevalence and spreading of diseases across populations. Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as Malaria, in ever greater detail.

\section{Hijacking Simulators}
\label{sec:hijackingsimulators}

\paragraph{Hijacking a simulator} describes the process by which a simulator's random number generators are replaced by calls to external sampling procedures, which are controlled by a probabilistic programming system (PPS). In practice, this amounts to performing a small number of surgical incisions into the simulator's source code in order to replace built-in calls to random number generators. E.g., given a simulator written in C++/Boost\cite{schaling2011boost}, a Gaussian distribution object \textit{boost::normal} is replaced with the corresponding \textit{pyprob\_cpp} distribution object, namely \textit{pyprob\_cpp::distributions::Normal}. Sampling from this distribution is then done by requesting the PPS to send a sample back to the simulator.

@@

% \begin{table}[h!]
% \footnotesize
% \begin{tabular}{p{3.6cm}|p{3.6cm}}
% Before  &  After  \\
% \midrule
% \begin{lstlisting}
% #include <gsl/gsl_cdf.h>
% #include <gsl/gsl_rng.h>
% #include <gsl/gsl_randist.h>
% //preamble
% double random::uniform_01 () {
%     double result =
% # ifdef 
% OM_RANDOM_USE_BOOST
%        rng_uniform01 ();
% # else
%  gsl_rng_uniform (rng.gsl_generator);
% # endif
%     return result;
%     }
% \end{lstlisting}&
% \begin{lstlisting}
% #include <pyprob_cpp.h>
% #include "xtensor/xadapt.hpp"
% // preamble
% double random::uniform_01 () {
% auto uniform = pyprob_cpp::distributions::Uniform(0,1);
% return pyprob_cpp::sample(uniform)(0);
% }
% \end{lstlisting}
% \end{tabular} 
% \end{table}
% As can be seen, writing the new sampling call is cleaner and syntactically simpler. 
%\bg{Add performance plot showing memory consumptions of pyprob and our new system.}

\section{Case Study: Ifakara (Tanzania)}
\label{sec:casestudy}

>>>>

\section{Introduction}
\label{sec:intro}
\bg{The audience is a probabilistic programming audience - so that is what we will target}

@@

\item Talk about a goal of probabilistic programming is to simultaneously simulate and perform inference on models conditioned on the observations. Although universal ppls have the ability to do this, they are constrained because the simulator has to be re-written as a program inside that system; this is both infeasible for sufficiently large simulators and inefficient. It would however be much more efficient to turn simulators into probabilistic programs... then talk briefly about the foundations created by baydin et al. (2 paragraphs)
\item Talk about the large body of simulators out there and briefly talk about the types of simulators
i.e population-based, event-based, single-object based and so fourth. {1 paragraph}
\item Talk, high level, about the notion of a posterior in terms of traces either add stuff from tom's thesis or directly cite. (1 paragraph)
\item Talk about the experiments we perform (high- level)

@@

\bg{This is not an intro - it is a collection of statements - will re-write later}

Simulators encode several years, if not decades, of research and development and so by their very nature are structurally complicated. Since the development happens over several decades, usually without documentation, understanding a code base written with legacy libraries and software is incredibly difficult and an almost impossible task for a non-expert, or expert who has not had access to the previous years of development. 
Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient} and extends the framework to encapsulate a more diverse range of simulators, in particular epidemiology simulators. Our framework provides a simple solution and makes it easy to transform arbitrary stochastic simulators into probabilistic programs, regardless of the complexity of the simulator and the language that the simulator is written in and for posterior inference. This enables two things. 1) It enables software developers and researchers to understand complicated code bases. 2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to inference outcome. This is critical in several domains, and is particularly important in the medical domain. We provide examples of both in Section~\textbf{TO ADD REF}.

@@

This paper first gives an overview of existing Malaria simulators (Section \ref{sec:background}), and proceed by introducing the necessary background on the \textit{pyprob} framework and the concept of universal probabilistic programming (Section \ref{sec:hijackingsimulators}). Our approach is then demonstrated and analysed in the context of a Malaria case study (Section \ref{sec:casestudy}).

\section{Background}
\label{sec:backgorund}

\begin{itemize}
\item Paragraph on Probabilistic programming
\item Paragraph on Simulators keep high-level focus on pop sims later
\item Paragraph on Baydin et al
\end{itemize}
In-silico simulators have become a crucial tool in evidence-based decision-making within a large number of disciplines, including statistical physics~\cite{landau_binder_2014}, financial modeling~\cite{jackel2002monte},
weather prediction~\cite{evensen1994sequential}, epidemiology~\cite{smith2008towards} and many others.
In many cases, simulation output can augment or even replace real data that may otherwise be costly or even impossible to generate.
Recent advances in hardware have enabled simulations to model increasingly complex systems.
Epidemiology studies the prevalence and spreading of diseases across populations. Recent advances in hardware have enabled simulations to model the dynamics of infectious diseases, such as Malaria, in ever greater detail.

% \section{Types of Models and Probabilistic Programming Languages}

% To bridge the gap between the statistics and machine learning community to the probabilistic programming community we provide a useful overview to expose our ideas more concretely.

% Probabilistic programming languages, broadly speaking\footnote{For a more comprehensive overview see \cite{rainforth2017automating}}, can be classfied into three categories, each type of langauge extends a particular model space, enabling an increasingly complex set models to be represented as programs, but in doing so the notion of posterior and what it means to do inference over the model, at least in a stastically traditional sense, can change.
% % \begin{figure}
% % \centering
% % \includegraphics[]{../models/types_of_pp.png}
% % \end{figure}
% The first type of langauges are called first-order languages where the priors are well defined and the \textit{joint} program density constructed via the program directly represent the unormalised stastical joint density $p(x,y)$
% \section{Methodology}
%  TODO

\section{Hijacking Simulators}

\paragraph{Hijacking a simulator} describes the process by which a simulator's random number generators are replaced by calls to external sampling procedures, which are controlled by a probabilistic programming system (PPS). In practice, this amounts to performing a small number of surgical incisions into the simulator's source code in order to replace built-in calls to random number generators. E.g., given a simulator written in C++/Boost\cite{schaling2011boost}, a Gaussian distribution object \textit{boost::normal} is replaced with the corresponding \textit{pyprob\_cpp} distribution object, namely \textit{pyprob\_cpp::distributions::Normal}. Sampling from this distribution is then done by requesting the PPS to send a sample back to the simulator. \bg{This will be extended}

 \label{sec:methods}

  \begin{algorithm}
   \caption{How to perform posterior inference over a stochastic simulator with 
   }
    \begin{algorithmic}[1]
      \Function{Convert Simulator to a Probabilistic Program}{$A$}\Comment{}

        \State replace raw samples statements $f_{t}(x_{t} | \eta_{t}) $
        \State Create thin wrapper
        \State Create Containerized Environment $C_{i}$
        \State Create Connection
        \State Run Program
       \EndFunction
       \Function{Create thin wrapper}{$f_{t}(x_{t} | \eta_{t})$}
       \State to add
       \EndFunction
       \Function{Create Containerized Environment }{$\Delta_{i}$}\Comment{Where $\Delta_{i}$ is one of $i$-th dependencies}
       \State to add
       \EndFunction

@@

@@

@@

% \begin{table}[h!]
% \footnotesize
% \begin{tabular}{p{3.6cm}|p{3.6cm}}
% Before  &  After  \\
% \midrule
% \begin{lstlisting}
% #include <gsl/gsl_cdf.h>
% #include <gsl/gsl_rng.h>
% #include <gsl/gsl_randist.h>
% //preamble
% double random::uniform_01 () {
%     double result =
% # ifdef 
% OM_RANDOM_USE_BOOST
%        rng_uniform01 ();
% # else
%  gsl_rng_uniform (rng.gsl_generator);
% # endif
%     return result;
%     }
% \end{lstlisting}&
% \begin{lstlisting}
% #include <pyprob_cpp.h>
% #include "xtensor/xadapt.hpp"
% // preamble
% double random::uniform_01 () {
% auto uniform = pyprob_cpp::distributions::Uniform(0,1);
% return pyprob_cpp::sample(uniform)(0);
% }
% \end{lstlisting}
% \end{tabular} 
% \end{table}
% As can be seen, writing the new sampling call is cleaner and syntactically simpler. 
%\bg{Add performance plot showing memory consumptions of pyprob and our new system.}
\subsection{Toy Model}
\subsection{Real-world Example}
\label{sec:casestudy}
